---
title: "Assignment 7"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

##The three input files in htm, json and xml formats are uploaded to Github
https://github.com/tponnada/DATA607/blob/master/convertcsv.htm, 
https://github.com/tponnada/DATA607/blob/master/convertcsv.json and 
https://github.com/tponnada/DATA607/blob/master/Rfile-2.xml

alongwith the original csv file - https://raw.githubusercontent.com/tponnada/DATA607/master/Rfile.csv

##The original table is loaded

```{r}
library(tidyverse)
library(RCurl)
library(XML)
library(rjson)
library(jsonlite)
library(DT)
library(plyr)
library(methods)
original <- read.csv("https://raw.githubusercontent.com/tponnada/DATA607/master/Rfile.csv", stringsAsFactors = FALSE)
datatable(original)
```


##Read data in first using an html link and create a corresponding data frame. I created the html file, uploaded files to Github and consequently downloaded them to local computer from where they are read.

```{r html}
htmlparsed <- htmlParse(file = "/Users/tponnada/Desktop/DATA607/Rfile.htm")
html_table <- readHTMLTable(htmlparsed, stringsAsFactors = FALSE)
html_table <- html_table[[1]]
html_books <- as.data.frame(html_table); html_books
```

##Read data in using an xml file and create a corresponding data frame.

```{r xml, echo=FALSE}
xml_parsed <- xmlParse(file = "/Users/tponnada/Downloads/Rfile-2.xml")
xml_books <- xmlToDataFrame(xml_parsed)
datatable(xml_books)
```

##Read data in using an JSON file and create a corresponding data frame.

```{r}
json_url <- "/Users/tponnada/Downloads/convertcsv.json"
json <- fromJSON(json_url); json
json_books <- as.data.frame(json); json_books
```

## Conclusion
View and compare the three separate data frames. The three observations are not equal. The json format is equal to the original csv file, the html file has a blank row and some additional spaces and the xml parse gives a lump of content.

```{r}
View(html_books)
View(xml_books)
View(json_books)
str(html_books)
str(xml_books)
str(json_books)
all.equal(original, html_books)
all.equal(original, json_books)
all.equal(original, xml_books)
```

